Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-30 paper/gd1.tex
fig4 caption

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Most stellar streams discovered around the Milky Way reside in the Galactic halo
and are therefore primarily sensitive to the dark matter distribution.
% The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
% to constrain the shape of the inner halo, and is a promising place to search for
% signatures of dark matter subhalo interactions.
Here we present a sample of highly-probable members of GD-1, the longest and most prominent thin stream in the Milky Way, selected to have
small parallaxes and retrograde proper motions using data from the \gaia\
mission, and \pans\ photometry consistent with an old and metal-poor population
at a distance of $\sim10~\kpc$.
This selection extends the stream by $20^\circ$, reveals a plausible location of
the progenitor, and shows significant density variations along the stream in
unprecedented detail.
In addition to confirming several prominent gaps detected in photometric maps of GD-1, for the first time we find evidence
of stream members off of the main track.
These features might be an indication of complex internal kinematics within a progenitor unlike any of the known globular clusters, or a result of a stream interaction with a massive perturber.
% Detailed modeling of the positional and velocity offsets
% or an indication of complex internal kinematics within the
% stream progenitor.
% Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
% If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
% If caused by a subhalo interaction, we estimate that the detected underdensity
% and the associated spur could have been induced by a perturbation of a
% $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{gd1_sample.pdf}
\end{center}
\caption{
On-sky positions of likely GD-1 members, in the coordinate system aligned with the stream.
GD-1 is apparent as an overdensity in retrograde proper motions (orange box on the top right panel), so a proper motion selection already reveals the stream in positions of individual stars (top left).
The stream population is older and more metal poor than the field Milky Way population, so the stream also stands out in the color-magnitude diagram (bottom right).
Additional photometric selection of the stream's main sequence (orange region in the CMD) unveils the stream in unprecedented detail (bottom left).
This map extends the stream by $20^\circ$, identifies a plausible location of the progenitor at $\phi_1\sim13^\circ$, confirms the existence of a significant gap at $\phi_1\sim-20^\circ$, shows additional under-densities at $\phi_1\sim-40^\circ,-5^\circ$ and for the first time finds stream overdensities off from the main track of a cold stream at $(\phi_1,\phi_2)\sim(-35^\circ,-1^\circ)$ and $(-15^\circ,-2^\circ)$.
\label{fig:selection}
}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{track_observables.pdf}
\end{center}
\caption{%
\todo{APW}
\label{fig:track-and-model}
}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{orbitfit.pdf}
\end{center}
\caption{%
\todo{APW}
\label{fig:orbitfit}
}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{features.pdf}
\end{center}
\caption{%
\todo{AB}
\label{fig:features}
}
\end{figure}

>>>>

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Most stellar streams discovered around the Milky Way reside in the Galactic halo
and are therefore primarily sensitive to the dark matter distribution.
% The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
% to constrain the shape of the inner halo, and is a promising place to search for
% signatures of dark matter subhalo interactions.
Here we present a sample of highly-probable members of the longest cold stream, GD-1, selected to have
small parallaxes and retrograde proper motions using data from the \gaia\
mission, and \pans\ photometry consistent with an old and metal-poor population
at a distance of $\sim10~\kpc$.
This selection extends the stream by $20^\circ$, reveals a plausible location of
the progenitor, and shows significant density variations along the stream in
unprecedented detail.
In addition to confirming several prominent gaps detected in photometric maps of GD-1, for the first time we find evidence
of stream members off of the main track.
These features might be an indication of complex internal kinematics within a progenitor unlike any of the known globular clusters, or a result of a stream interaction with a massive perturber.
% Detailed modeling of the positional and velocity offsets
% or an indication of complex internal kinematics within the
% stream progenitor.
% Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
% If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
% If caused by a subhalo interaction, we estimate that the detected underdensity
% and the associated spur could have been induced by a perturbation of a
% $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{gd1_sample.pdf}
\end{center}
\caption{
On-sky positions of likely GD-1 members, in the coordinate system aligned with the stream.
GD-1 is apparent as an overdensity in retrograde proper motions (orange box on the top right panel), so a proper motion selection already reveals the stream in positions of individual stars (top left).
The stream population is older and more metal poor than the field Milky Way population, so the stream also stands out in the color-magnitude diagram (bottom right).
Additional photometric selection of the stream's main sequence (orange region in the CMD) unveils the stream in unprecedented detail (bottom left).
This map extends the stream by $20^\circ$, identifies a plausible location of the progenitor at $\phi_1\sim13^\circ$, confirms the existence of a significant gap at $\phi_1\sim-20^\circ$, shows additional under-densities at $\phi_1\sim-40^\circ,-5^\circ$ and for the first time finds stream overdensities off from the main track of a cold stream at $(\phi_1,\phi_2)\sim(-35^\circ,-1^\circ)$ and $(-15^\circ,-2^\circ)$.
}
\label{fig:selection}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{track_observables.pdf}
\end{center}
\caption{%
\todo{APW}
}
\label{fig:track-and-model}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{orbitfit.pdf}
\end{center}
\caption{%
\todo{APW}
}
\label{fig:orbitfit}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{features.pdf}
\end{center}
\caption{
Color-magnitude and proper-motion diagrams of fields associated with a spur above GD-1 (top), and fields related to the blob below GD-1 (bottom).
The left-most pair of panels show properties in stream fields (darkest rectangles in the middle panel), the middle pairs present spur and blob (medium rectangles), while the right-most panels have fields on the opposite side from spur and blob (lightest rectangles).
Selection boxes used to select stars shown in the middle panel are shown in orange in all CMD and proper-motion panels.
Both the spur and the blob fields have more stars in the CMD selection box than the corresponding control fields.
Proper motions of these stars follow the distribution of stars in adjacent stream fields, thus confirming the association of these off-track features with GD-1.
}
\label{fig:features}
\end{figure}
